---
title: "NYC Citi Bikes Notes"
output: html_notebook
---

```{r}
library(tidyverse)
library(tsibbledata)
library(tsibble)
library(lubridate)
library(jsonlite)
library(leaflet)
library(sf)
library(e1071)
```

```{r}
bikes_data <- tsibbledata::nyc_bikes


glimpse(bikes_data)
```

What is the pattern of bike hires over time
  - year
  - month
  - week
  - day
  
```{r}
bikes_data <- bikes_data %>% 
  mutate(year = year(start_time),
         month = month(start_time, label = TRUE, abbr = FALSE),
         day = wday(start_time, label = TRUE, abbr = FALSE),
         day_of_month = mday(start_time),
         start_hod = strftime(round_date(bikes_data$start_time, "30 minutes"),
                              format= "%H:%M",
                              tz = "America/New_York"),
         end_hod = strftime(round_date(bikes_data$stop_time, "30 minutes"), 
                              format= "%H:%M",
                              tz = "America/New_York"))



# 
# bikes_data %>% 
#   mutate(start_hod = strftime(bikes_data$start_time, 
#                               format= "%H:%M",
#                               tz = "America/New_York"))
  
bikes_data %>% 
  distinct(year)
# Only shows 2018

bikes_data %>% 
  ggplot(aes(x = fct_rev(month)))+
  geom_histogram(stat = "count", fill = "#6B7FBA")+
  labs(
    x = "Months",
    y = "Total Bike Rentals",
    title = "Bike Rentals by Month for 2018"
      )+
  coord_flip()+
  theme_classic()

  

bikes_data %>% 
  index_by(month) %>% 
  group_by(bike_id) %>% 
  count()


bikes_data %>% 
  ggplot(aes(x = fct_shift(fct_rev(day), -1)))+
  geom_histogram(stat = "count", fill = "#3DA4CA")+
  labs(
    x = "Months",
    y = "Total Bike Rentals",
    title = "Bike Rentals by Day for 2018"
      )+
  coord_flip()+
  theme_classic()


```
  
 Calculating Ride Duration
 
```{r}
bikes_data <- bikes_data %>% 
  mutate(ride_duration = as.numeric(as.duration(stop_time - start_time), "minutes")) %>% 
  mutate(ride_duration = round(ride_duration, digits = 2))

 # Average duration of bike rides by month
bikes_data %>% 
  filter(type == "Subscriber") %>% 
  index_by(month) %>% 
  summarise(avg_ride_duration = mean(ride_duration)) %>% 
  ggplot(aes(x = fct_rev(month), y = avg_ride_duration))+
  geom_col()+
  labs(
    x = "Month",
    y = "Average Duration of Hire (minutes)",
    title = "Average duration of bike hires by month"
  )+
  coord_flip()+
  theme_classic()

bikes_data %>% 
  filter(type == "Subscriber") %>% 
  filter(ride_duration <= 45.00) %>% 
  index_by(month) %>% 
  summarise(avg_ride_duration = mean(ride_duration)) %>% 
  ggplot(aes(x = fct_rev(month), y = avg_ride_duration))+
  geom_col()+
  labs(
    x = "Month",
    y = "Average Duration of Hire (minutes)",
    title = "Average duration of bikes hired by month"
  )+
  coord_flip()+
  theme_classic()

```

### Add Rider age and check if it impacts type of user

```{r}
bikes_data <- bikes_data %>% 
  mutate(rider_age = year - birth_year)

bikes_data %>% 
  filter(rider_age <= 60) %>% 
  mutate(avg_age = mean(rider_age)) %>% 
  mutate(avg_age = round(avg_age, digits = 1)) %>% 
  distinct(avg_age)

# 36 Average Age

bikes_data %>% 
  filter(type == "Subscriber") %>% 
  index_by(month) %>% 
  summarise(avg_age = mean(rider_age)) %>% 
  mutate(avg_age = round(avg_age, digits = 1)) %>% 
  ggplot(aes(x = month, y = avg_age))+
  geom_col()

bikes_data %>% 
  filter(type == "Customer") %>% 
  index_by(month) %>% 
  summarise(avg_age = mean(rider_age)) %>% 
  mutate(avg_age = round(avg_age, digits = 1)) %>% 
  ggplot(aes(x = month, y = avg_age))+
  geom_col()
```


```{r, warning=FALSE, message = FALSE}
bikes_data %>% 
  filter(type == "Subscriber") %>% 
  ggplot(aes(x = rider_age))+
  geom_histogram(colour = "white", fill = "#6B7FBA")+
  geom_vline(xintercept = median(bikes_data$rider_age),
             col = "firebrick",
             lwd = 2) +
  annotate("text",
           x = median(bikes_data$rider_age) * 1.3,
           y = median(bikes_data$rider_age) * 18,
           label = paste("Average Rider Age = ", median(bikes_data$rider_age)),
           col = "firebrick",
           size = 4
           )+
  labs(
    x = "Age of User",
    y = "Number of Users"
  )+
  theme_classic()+
  xlim(20, 75)

```

### Adding potential rates for customer journeys

```{r}

bikes_data %>% 
  filter(type == "Customer") %>% 
  mutate(ride_fee = ifelse(ride_duration >= 45, 15, 3.99)) %>% 
  ggplot(aes(x = month, y = ride_fee))+
  geom_col()+
  labs(
    x= "Months",
    y = "Estimated revenue ($)",
    title = "Estimated Revenue from non-subscribers in 2018"
  )
  

```

###  Comparing gender distribution for usage
 
```{r}
bikes_data %>% 
  ggplot(aes(x = fct_rev(month), fill = gender))+
  geom_histogram(stat = "count", position = "dodge")+
  labs(
    x = "Months",
    y = "Number of users",
    title = "Number of recorded users by gender"
  )+
  coord_flip()


bikes_data %>%
  group_by(gender) %>% 
  count()

3069 / nrow(bikes_data) 

```
 
### Plot locations of bike stations on a map 

```{r}

# bikes_data %>%
#   filter(start_station %in% c("3186", "3203", "3195", "3183", "3202")) %>%
#   filter(end_station %in% c("3186", "3183", "3203", "3195", "3202")) %>%
#   leaflet() %>%
#   addTiles() %>%
#   addMarkers(
#     lat = ~unique(start_lat),
#     lng = ~unique(start_long),
#     popup = ~unique(start_station),
#     clusterOptions = markerClusterOptions()
#     )%>%
#     addMarkers(
#     lat = ~unique(end_lat),
#     lng = ~unique(end_long),
#     popup = ~unique(end_station),
#     clusterOptions = markerClusterOptions()
#     )


bikes_data %>% 
  filter(start_hod <= "10:00") %>% 
  count(start_station) %>% 
  arrange(desc(n))

bikes_data %>%
  filter(
    start_station %in% c("3203", "3213", "3267", "3269", "3186"),
    end_station %in% c("3186", "3183", "3195", "3202", "3639")) %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(
    color = "blue",
    lat = ~unique(end_lat),
    lng = ~unique(end_long),
    popup = ~unique(end_station)
  )%>%
  addCircleMarkers(
    color = "green",
    lat = ~unique(start_lat),
    lng = ~unique(start_long),
    popup = ~unique(start_station)
  )
#   Start Stations

# 3203 is residential area
# 3213 is residential area
# 3267 is residential area
# 3269 is residential area
# 3186 is residential area

#   End Stations

# 3186 retail/commercial/finance hub
# 3183 right next to Ferry Terminal
# 3195 right next to transport hub (trains & buses)
# 3202 - uni/mall area
# 3639 financial sector

bikes_data %>% 
  count(bike_id, start_station) %>% 
  arrange(desc(n))

```

### Calculating Distance

No point in calculating distance as start/end stations are all next to each other and no distance tracker data is included


### Checking to see if standard commuting times apply

```{r}

# bikes_data %>%
#  # filter(month == "January") %>% 
#   ggplot(aes(x = start_hod))+
#   geom_histogram(stat = "count")+
#   labs(
#     x = "Journey Start Time"
#   )+
#   facet_wrap(~month)

  bikes_data %>% 
    ggplot(aes(x = end_hod, group = type, colour = type))+
    geom_line(stat = "count")+
    labs(
      x = "Journey End Times",
      y = "Number of Journeys",
      title = "Journey times by user type"
    )+
    guides(colour = guide_legend(title = "User Type")
    )+
    scale_x_discrete(
      breaks = c("00:00","03:00", "06:00", "09:00", "12:00", "15:00", "18:00", "21:00"))
  
  bikes_data %>% 
    ggplot(aes(x = start_hod, group = type, colour = type))+
    geom_line(stat = "count")+
    labs(
      x = "Journey Start Times",
      y = "Number of Journeys",
      title = "Journey times by user type"
    )+
    guides(colour = guide_legend(title = "User Type")
    )+
    scale_x_discrete(
      breaks = c("00:00","03:00", "06:00", "09:00", "12:00", "15:00", "18:00", "21:00"))


```
