---
title: "R Notebook"
output: html_notebook
---
# Libraries
```{r}
library(readr)
library(here)
library(tidyverse)
```

# Read in the relevant files
```{r}
candy_2015 <- read_csv(here::here("clean_data/candy_2015.csv"))
candy_2016 <- read_csv(here::here("clean_data/candy_2016.csv"))
candy_2017 <- read_csv(here::here("clean_data/candy_2017.csv"))

```

# Combine all dataframes into one hideous frame

```{r}
candy_master <- bind_rows(candy_2015, candy_2016, candy_2017)
names(candy_master)
# Looking at names there should be some manual renaming for easier use

candy_master <- candy_master %>% 
  rename(
    pb_kisses = 
      anonymous_brown_globs_that_come_in_black_and_orange_wrappers,
    full_size_candy = any_full_sized_candy_bar,
    other_brach_products = brach_products_not_including_candy_corn,
    restaurant_free_candy = 
      candy_that_is_clearly_just_the_stuff_given_out_for_free_at_restaurants,
    currency = cash_or_other_forms_of_legal_tender,
    toblerone = tolberone_something_or_other,
    chick_o_stick = chick_o_sticks_we_don_t_know_what_that_is,
    mm_circ_peanututs = those_odd_marshmallow_circus_peanut_things,
    poi_season3 = 
      person_of_interest_season_3_dvd_box_set_not_including_disc_4_with_hilarious_outtakes,
    sourpatch_kids = sourpatch_kids_i_e_abominations_of_nature,
    vials_of_syrup = 
      vials_of_pure_high_fructose_corn_syrup_for_main_lining_into_your_vein
    )

names(candy_master)

# Relocate country so its near the start of the data frame
candy_master <- candy_master %>% 
  relocate(country, .after = age) %>% 
  relocate(gender, .after = age) %>% 
  select(-state)
    
```

# Tackling country column

```{r}
candy_master %>% 
  select(year, country, state)

```

# Initally looking at the data, 2015 has no data for country or state
 Initially while sorting through country/state filtering out all 2015 
 entries will be easiest. Using state can be helpful if country is NA 
 but state isn't. Once all country entries have been sorted we can drop 
 state altogether as it isn't relevant to any questions
 
```{r}

# Use filter and select to create a string we can use to uniform all entries
candy_master %>% 
  distinct(country)
```
 
# Create some variables that we can use for organizing entries, these can be 
  used for multiple years and we can add as we check other years entries

```{r}

pattern_usa <- c(
      "usa", "US", "USA", "United States of America", "uSA", "united states", 
      "United States", "us", "USSA", "U.S.A", "Murica", "USA!", "Usa", "U.S.", 
      "Us", "America", "Units States", "United states", "USA USA USA", 
      "the best one - usa", "USA! USA! USA!", "u.s.", 
      "The Yoo Ess of Aaayyyyyy", "united states of america", "USA!!!!!!", 
      "USA! USA!", "United Sates", "Sub-Canadian North America... 'Merica", 
      "Trumpistan", "U.s.", "Merica", "UNited States", "United Stetes", 
      "america", "USA USA USA USA", "United State", "god's country",
      "USA (I think but it's an election year so who can really tell)", "U.S.A.",
      "Unhinged states","USAUSAUSA", "United staes", "u.s.a.", "USAUSAUSA", 
      "United States of America", "US of A", "unhinged states", "Unites States",
      "The United States", "North Carolina", "Unied States", "U S", 
      "The United States of America", "unite states", "USA? Hard to tell anymore..",
      "'merica", "usas", "Pittsburgh", "New York", "California", "USa", 
      "I pretend to be from Canada, but I am really from the United States.", 
      "United Stated", "Ahem....Amerca", "New Jersey", "United ststes", 
      "United Statss", "murrika", "USAA", "Alaska", "united States", "u s a",
      "United Statea", "united ststes", "USA USA USA!!!!", "EUA", "United States of America"
)

pattern_uk <- c("UK", "england", "uk", "United Kingdom", "England","United Kindom",
                "U.K.", "Uk", "Scotland", "endland", "United kingdom")

pattern_canada <- c("Canada", "canada", "Canada`", "CANADA", "soviet canuckistan"
                    )

pattern_other <- c("Japan", "france", "A tropical island south of the equator", 
                   "Switzerland", "Neverland", "this one", "Korea","belgium", 
                   "croatia", "Portugal", "españa", "Panama",
                   "there isn't one for old men", "one of the best ones", "France",
                   "Australia", "hungary", "Austria", "Somewhere", "New Zealand",
                   "Germany", "Mexico", "Brasil", "South Korea", 
                   "Philippines", "sweden", "The Netherlands", "Finland", "China",
                   "germany", "See above", "kenya", "Netherlands", "Not the USA or Canada",
                   "netherlands", "Denial", "UAE", "finland", "Europe", "Earth", 
                   "Costa Rica", "insanity lately","The republic of Cascadia", 
                    "Cascadia", "cascadia", "Greece", "australia", "A", "Can",
                   "Canae", "Ireland", "South africa", "Iceland", "Denmark", "UD",
                   "Indonesia", "Atlantis", "N. America", "Singapore", "Taiwan", 
                   "hong kong", "spain", "Sweden", "Hong Kong", "Narnia", 
                   "subscribe to dm4uz3 on youtube", "I don't know anymore",
                   "Fear and Loathing", "46", "45.0", "45", "44.0", "35", "32",
                   "30.0", "1")

#Added the ages to the other column, if I have time i would like to change them
# to countries depending on state entered
```

# Use created variables to filter and organise the country column
```{r}

candy_master$country[candy_master$country %in% pattern_usa] <- "USA"
candy_master$country[candy_master$country %in% pattern_uk] <- "UK"
candy_master$country[candy_master$country %in% pattern_canada] <- "Canada"
candy_master$country[candy_master$country %in% pattern_other] <- "Other"
# TO DO: Expand filters for all years
#        Save variables into seperate notepad doc and write them in

candy_master %>% 
  group_by(country) %>% 
  count() %>% 
  arrange(desc(country))

```
 
 
# Sorting the age column

```{r}
#remove the decimal points and change to numeric to remove characters etc
candy_master <- candy_master %>% 
  mutate(age = str_replace(age, "\\..*", "")) %>%
  mutate(age = as.numeric(age)) %>%
  arrange(desc(age))

# This will remove at least one entry that had an age plus some other words,
# if time return and modify

#replace all values that are unlikely with 0
candy_master$age[candy_master$age > 120] <- 0

#Check it worked
candy_master %>% 
  distinct(age)



```

## 1.4.2 Analysis questions

# 1 What is the total number of candy ratings given across the three years. 
(Number of candy ratings, not the number of raters. Don’t count missing values)

```{r}

# So we need to count each candy column and how many entries each has that aren't
# NA value then combine each columns total, would this include the non candy
# answers like the DVD's?
names(candy_master)

q1_candy_ratings <- candy_master %>%
  select(6:128)#select the columns that are to do with ratings and save them

# Initial plan was to count the na's then sub from the total

sum(is.na(q1_candy_ratings)) # total NA's 396711

# Can simply just count how many entries there are that are not NA
total_candy_ratings <- sum(!is.na(q1_candy_ratings))

# 771,914 Candy Ratings over the years

```

# 2 What was the average age of people who are going out trick or treating?

```{r}

candy_master %>% 
  filter(trick_or_treating == "Yes") %>% 
  mutate(average_age = mean(age, na.rm = TRUE)) %>% 
  select(average_age)

#average age was 33.6

```

# 3 What was the average age of people who are not going trick or treating?

```{r}
candy_master %>% 
  filter(trick_or_treating == "No") %>% 
  mutate(average_age = mean(age, na.rm = TRUE)) %>% 
  select(average_age)

# Average age 37.9
```

# 4 For each of joy, despair and meh, which candy bar received the most of these ratings?

```{r}

q4_candy <- candy_master %>% 
  select(6:128)

q4_candy %>% 
 summarise(across(everything(), .fns = ~str_count(., pattern = "JOY"))) %>% 
  summarise(across(everything(), .fns = ~sum(., na.rm = TRUE))) %>% 
      pivot_longer(
        cols = everything(),
        names_to = "candy_type",
        values_to = "rating") %>% 
  arrange(desc(rating))

# Full Size Candy Bars received the most Joy ratings
  
q4_candy %>% 
 summarise(across(everything(), .fns = ~str_count(., pattern = "DESPAIR"))) %>% 
  summarise(across(everything(), .fns = ~sum(., na.rm = TRUE)))  %>%  
  pivot_longer(
    cols = everything(),
    names_to = "candy_type",
    values_to = "rating") %>%
  arrange(desc(rating))

# Broken Glow Sticks received the most despair ratings

q4_candy %>% 
 summarise(across(everything(), .fns = ~str_count(., pattern = "MEH"))) %>% 
  summarise(across(everything(), .fns = ~sum(., na.rm = TRUE))) %>% 
    pivot_longer(
    cols = everything(),
    names_to = "candy_type",
    values_to = "rating") %>% 
  arrange(desc(rating))

# Lollipops received the most meh ratings

```

#Q5 How many people rated Starburst as despair?

```{r}
# Using method from previous answer and filtering we can easily find this
q4_candy %>% 
 summarise(across(everything(), .fns = ~str_count(., pattern = "DESPAIR"))) %>% 
  summarise(across(everything(), .fns = ~sum(., na.rm = TRUE)))  %>%  
  pivot_longer(
    cols = everything(),
    names_to = "candy_type",
    values_to = "rating") %>%
  filter(candy_type == "starburst")
  
# 1990 people voted starburst with despair
```

