---
title: "R Notebook"
output: html_notebook
---

```{r}
library(readr)
library(readxl)
library(here)
library(tidyverse)
library(rebus)

bird_data <- read_csv(here::here("clean_data/clean_bird_data.csv"))
ship_data <- read_csv(here::here("clean_data/clean_ship_data.csv"))

```

# Now that both clean data files are read in we can join them for initial view

```{r}
seabirds_data <- left_join(bird_data, ship_data, by = c("record_id" = "record_id"))


```

```{r}

names(seabirds_data) 
# note that record.x and record.y have been created 
# from the join

head(seabirds_data, 10) 
# record y can be dropped as it is not a unique 
# modifier for the table. 
# record.x can be renamed to record

seabirds_data <- seabirds_data %>% 
  rename("record" = record.x) %>% 
  select(!record.y)

seabirds_data %>% 
    summarise(across(.fns = ~sum(is.na(.))))

seabirds_data %>% 
  filter(is.na(lat))


```

# Looking at questions


    Which bird had the most individual sightings?
    Which bird had the highest total count?
    Which bird had the highest total count above a latitude of -30?
    How many different types of birds were only ever seen in groups of 1?
    How many penguins were seen? (Hint: there are many types of penguin)


#Q1 Which bird had the most individual sightings? 

```{r}
seabirds_data %>% 
  group_by(species_common_name) %>% 
  summarise(species_common_name, #get our required columns for the answer
            species_scientific_name, 
            species_abbreviation, 
            total_sightings = n()) %>% # count in the groups for individual 
                                       # sightings
  distinct() %>% 
  arrange(desc(total_sightings)) 

# Wandering albatross sensu lato PL2 (Plummage stage)
# Wandering albatross of various plummage stages very common
```

#Q2 Which bird had the highest total count?

```{r}
seabirds_data %>% 
  group_by(species_common_name) %>% 
  summarise(species_common_name, #get required columns for the answer
            species_scientific_name, 
         species_abbreviation, 
         total_count = sum(count))%>%
  distinct() %>% #filter to distinct names so we can see all the top results
  arrange(desc(total_count))

# Short-tailed Shearwater, 982553
```

#Q3 Which bird had the highest total count above a latitude of -30?

```{r}
seabirds_data %>% 
  filter(lat > -30) %>%  # filter latitude column to only show counts higher than -30
  group_by(species_common_name) %>% #group by bird type for accurate count
  summarise(species_common_name, #get only the required columns for the answer
            species_scientific_name, 
            species_abbreviation, 
            total_count = sum(count))%>% 
  distinct() %>% #as we are doing a count we can filter to distinct names
  arrange(desc(total_count))

# Wedge tailed Shearwater
```
#Q4  How many different types of birds were only ever seen in groups of 1?

```{r}
seabirds_data %>% 
  filter(count == 1) %>%  # filter the count column to only show rows with a 
  # single count
  distinct(species_common_name) %>% #show only distinct names in the column
  count() # count how many are left after filter

# 298 different types of birds

```
#Q5 How many penguins were seen? (Hint: there are many types of penguin)
```{r}

seabirds_data %>% 
  str_detect("enguin") # Check to see if "enguin" will be enough for a catch all

seabirds_data %>% 
 filter(
   grepl("enguin", species_common_name)) %>%  # filter by chosen word, not using
  #regex as wasn't necessary
  summarise(total_penguins = sum(count)) # get total count of remaining 
 # rows count value

#158 penguins
```

